---
title: "Homework 3"
author: Student 1 and Student 2
output: pdf_document
date: "2024-11-5"
font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.table("hcmv.txt",  header = TRUE)
```

# 0. Contribution Statement

## Student 2

Student 2 mainly worked on questions 2 and 5 and provided graphs for question 4.

\pagebreak

\pagenumbering{arabic}

# Introduction

### Data

This data is from a publication by Chee et al. and describes the DNA sequence of CMV. A CMV DNA molecule has 229,354 complementary pairs of letters or base pairs. These scientists are in search of special patterns in the virus’ DNA that contains instructions for its reproduction: replication. They discovered a total of 296 palindromic sequences, each of which are at least 10 pairs long.

### Objective

The goals of this report are to investigate the distribution of genes among this DNA sequence.

\pagebreak

# Basic Analysis

## Question 1: RANDOM SCATTER

### Methods

We used ```replicate``` to take 100 simulations of a DNA sequence, with each simulation containing 296 palindrome sites. 

```{r data, echo=FALSE}
# SIMPLIFY = FALSE KEEPS EACH RESULT AS A SEPARATE VECTOR
n = 100
sequence_len = 229354
palindrome_sites = 296
set.seed(42)
simulations <- replicate(n, sample(1:sequence_len, palindrome_sites, replace = FALSE), simplify = FALSE)
```

QUANTITATIVE ANALYSIS:

We will calculate the average distances between palindrome sites for each simulation to determine whether the palindrome sites are randomly distributed or if there are patterns or clusters. 

```{r}
# For each simulation, calculate the differences between the data points (palindrome sites), and take the mean
average_distances <- sapply(simulations, function(simulation){
    sorted_simulation <- sort(simulation) # Sort simulation to get correct distances
    distances <- diff(sorted_simulation)
    mean(distances)
})
```

Now, we have a list of the average distances between palindrome sites in a DNA sequence. We will compute the mean of the average distances between the palindrome sites to provide a central measure of approximation.

```{r}
# On average, palindrome sites tend to be about 771.755 bases apart in simulated datasets. 
mean(average_distances)
```

QUALITATIVE ANALYSIS:
```{r}
histogram <- hist(average_distances, breaks = 10, col = 'lightblue', main = 'Differences in Simulated Palindrome Distances', xlab = 'Average Distance Between Palindromes')
```
[TYPE ANALYSIS OF HISTOGRAM HERE]

```{r}
image(matrix(average_distances, ncol=1), col=heat.colors(max(average_distances)), main="Heatmap of Simulated Palindrome Density")
```
[TYPE ANALYSIS OF HEATMAP HERE]

### Analysis
[KS ANALYSIS]
[CHECK CONDITIONS]
[CONDUCT TEST]

### Conclusion

# Question 2