---
title: "Homework 3"
author: Heidi Tam and Student 2
output: pdf_document
date: "2024-11-5"
font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.table("hcmv.txt",  header = TRUE)
```

# 0. Contribution Statement

## Student 2

Student 2 mainly worked on questions 2 and 5 and provided graphs for question 4.

\pagebreak

\pagenumbering{arabic}

# Introduction

### Data

This data is from a publication by Chee et al. and describes the DNA sequence of CMV. A CMV DNA molecule has 229,354 complementary pairs of letters or base pairs. These scientists are in search of special patterns in the virus’ DNA that contains instructions for its reproduction: replication. They discovered a total of 296 palindromic sequences, each of which are at least 10 pairs long.

### Objective

The goals of this report are to investigate the distribution of genes among this DNA sequence.

\pagebreak

# Basic Analysis

## Question 1

### Methods

```{r data, echo=FALSE}

```

### Analysis
### Conclusion

# Question 2