---
title: "Homework 3"
author: Student 1 and Student 2
output: pdf_document
date: "2024-11-5"
font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.table("hcmv.txt",  header = TRUE)
```

# 0. Contribution Statement

## Student 2

Student 2 mainly worked on questions 2 and 5 and provided graphs for question 4.

\pagebreak

\pagenumbering{arabic}

# Introduction

### Data

This data is from a publication by Chee et al. and describes the DNA sequence of CMV. A CMV DNA molecule has 229,354 complementary pairs of letters or base pairs. These scientists are in search of special patterns in the virus’ DNA that contains instructions for its reproduction: replication. They discovered a total of 296 palindromic sequences, each of which are at least 10 pairs long.

### Objective

The goals of this report are to investigate the distribution of genes among this DNA sequence.

\pagebreak

# Basic Analysis

## Question 1: RANDOM SCATTER

### Methods

Here, we are using ```sample.int``` to conduct 1 simulation assuming a uniform spread. 
```{r}
N = 229354
n = 296
set.seed(10)
random_1 <- as.vector(sample.int(N, size=n, replace=FALSE)) # locations uniformly randomly generated
```

We're interested in seeing how simulations vary when repeated many times, so we used ```replicate``` to take 100 simulations of a DNA sequence, with each simulation containing 296 palindrome sites. 

```{r data, echo=FALSE}
# SIMPLIFY = FALSE KEEPS EACH RESULT AS A SEPARATE VECTOR
num_sim = 100
N = 229354
n = 296
set.seed(10)
simulations <- replicate(n, sample(1:N, n, replace = FALSE), simplify = FALSE)
```

QUANTITATIVE ANALYSIS:

We will calculate the average distances between palindrome sites for each simulation to determine whether the palindrome sites are randomly distributed or if there are patterns or clusters. 

```{r}
# For each simulation, we will calculate the differences between the data points (palindrome sites), and take the mean
average_distances <- sapply(simulations, function(simulation){
    sorted_simulation <- sort(simulation) # We sorted the simulation to get correct distances
    distances <- diff(sorted_simulation)
    mean(distances)
})
```

Now, we have a list of the average distances between palindrome sites in a DNA sequence. We will compute the mean and variance of the average distances between the palindrome sites to provide a central measure of approximation.

```{r}
# On average, palindrome sites tend to be about 771.755 bases apart in simulated datasets. 
mean(average_distances)
# The variance of the average distances, about 14.677, tells us there is moderate variability in the average distances between palindromes. We can confirm this with a qualitative analysis, as follows. 
var(average_distances)
```

QUALITATIVE ANALYSIS:
```{r}
histogram <- hist(average_distances, breaks = 10, col = 'lightblue', main = 'Differences in Simulated Palindrome Distances', xlab = 'Average Distance Between Palindromes')
```
The histogram of the average distance between palindromes is left-skewed, which suggests few small gaps between palindromes sites. The average distances appear to mostly be clustered between 770 and 780, but we also have palindromes distances that were as low as 755. This variation can indicate that the palindromes may not be uniformly spread and have clustering at particular locations, which was confirmed by the quantitative analysis above. This histogram is also unimodal, suggesting most palindrome sites are mostly spaced apart within a certain range. 

```{r}
image(matrix(average_distances, ncol=1), col=heat.colors(max(average_distances)), main="Heatmap of Simulated Palindrome Density")
```
We can visualize the spread of distances between palindromes in a heat map. The palindromes are generally uniformly spread out but experience clustering in certain areas. 

### Analysis
We will conduct a **KS Analysis** to test whether the distribution of palindrome sites in our simulated dataset is statistically similar to the real dataset. 
**H0:** The distribution of simulated distances and real distances are the same.
**H1:**: The distribution of simulated and real distances differ. 
We meet the conditions to perform this test because the samples are independent and the data is continuous.
We don't have the "real data", so we will take a sample and treat it like the "real data". 
```{r}
real_data_distances <- diff(sort(sample(1:N, n)))
# KS test
ks_test <- ks.test(average_distances, real_data_distances)
```
### Conclusion
With a p-value of p < 2.2 * 10^(-16), we can reject the null hypothesis, and there is a significant difference between the random scatters and real data. However, our inferences in the accuracy of this conclusion are limited because we simply drew another sample and treated it as the "real data", but our "real data", may not necessarily look like this. 

# Question 2: LOCATIONS & SPACING

# Question 3: COUNTS
### Methods
We will make our list of all possible DNA locations, ranging from 1 to the 229354, denoted by N.
```{r}
DNA <- 1:N
# INVESTIGATE REGIONS OF LENGTH 1000
split_DNA <- split(DNA, ceiling(seq_along(DNA) / 1000))
# function to count number of palindromes per region (observed counts)
count_per_region <- function(region, palindrome_locations){
  sum(region %in% data$location)
}
counts_1000 <- sapply(split_DNA, count_per_region, data$location)
```

### Analysis
REGIONS OF SIZE 1,000
We will conduct a chi-squared test, goodness of fit to compare the number of palindromes when each region has 1,000 bases.
**H0:** There is no difference between the number of palindromes in regions of length 1,000 and number of palindromes that we would expect from uniform random scatter.
**H1:** There is a difference between the number of palindromes in regions of length 1,000 and the number of palindromes we would expect from a uniform random scatter.

First, we will check the conditions to make sure we can conduct this test.
**Observed Counts**: Yes, we have observed counts of the number of palindromes. 
**Independence**: The counts of the palindromes in each chunk are independent since they are non-overlapping.
**Expected Frequency > 5**: When we have regions of length 1000 each, we would expect each region to have 296 / 230 chunks = 1.287 palindromes per chunk. We will still perform the chi-squared test, but it may not be accurate, so we need to be careful. 
**

``` {r}
# number of palindromes we would expect for each region
expected <- 296 / 230 # total palindromes / # chunks
expected_vec <- rep(expected, 230)
expected_prob <- expected_vec / sum(expected_vec)

# perform a chi-squared test to compare observed counts with expected values
chi_squared_test <- chisq.test(counts_1000, p = expected_prob)
```

With a p-value of 0.004 < 0.05, we reject the null hypothesis, and there is a difference between the number of palindromes in regions of length 1000 and a uniform random scatter. 

Now, we will repeat this process but with larger region sizes.
REGIONS OF SIZE 5,000
```{r}
split_DNA_5k <- split(DNA, ceiling(seq_along(DNA) / 5000))
counts_5000 <- sapply(split_DNA_5k, count_per_region, data$location)
expected_5k <- 296 / 46 # total palindromes / # chunks
```
Now that we have expected counts of 296 / 46 = 6.435, which is > 5, we meet the condition of having E_i >= 5, for all i. The rest of the conditions are still satisfied, as described above. 
```{r}
expected_vec_5k <- rep(expected_5k, 46)
expected_prob_5k <- expected_vec_5k / sum(expected_vec_5k)
chi_squared_test_5k <- chisq.test(counts_5000, p = expected_prob_5k)
```
Our p-value from this chi-squared test is now 0.02, which is < 0.05. We still reject the null hypothesis, and there is a difference between the number of palindromes in regions of length 1000 and a uniform random scatter. We will conduct this statistical test one last time with regions of size 10,000. 
REGIONS OF SIZE 10,000
```{r}
split_DNA_10k <- split(DNA, ceiling(seq_along(DNA) / 10000))
counts_10000 <- sapply(split_DNA_10k, count_per_region, data$location)
expected_10k <- 296 / 23 # All conditions for the chi-squared test have been satisfied, described previously.
expected_vec_10k <- rep(expected_10k, 23)
expected_prob_10k <- expected_vec_10k / sum(expected_vec_10k)
chi_squared_test_10k <- chisq.test(counts_10000, p = expected_prob_10k)
```
Our p-value went up once again! With a p-value of 0.317, p > 0.05. We fail to reject the null hypothesis, and there is no significant difference between the counts of palindromes in our non-overlapping regions of equal length with the counts we would expect from a uniform random scatter. 

### Conclusion
When our region of non-overlapping DNA is shorter, we are sampling less DNA and are more likely to think there is a significant difference of counts of palindromes in our DNA sequence compared to a uniform random scatter. As we increased our DNA region length, our p-value also went up. As we observe more DNA consecutively, we see that the pattern of palindromes does not really occur at such a high rate as it appears to in the smaller regions. A good analogy is that you are more likely to find palindromes in shorter words (eg 'mom', 'dad') than a full sentence (eg "Was it a car or a cat I saw?"). 